---
title: "SI618f15_hw1_report_andriesd.rmd" 
output: html_document
---
## Step 1: Load Data
First, the provided TSV data file is loaded into R using the **read.table()** function. Here are the first 15 rows of the data frame:

```{r, echo=FALSE}
library(ggplot2)
data = read.table("~/Desktop/SI601_618/SI618/Week1/countrydata_withregion.tsv", header=TRUE, sep="\t", quote = "", fill=TRUE)
head(data, 15)
```

## Step 2: Scatter plot of log transformed data
Natural logarithms of the area and the population of each country are computed and used to produce the following scatter plot using the **qplot()** function.

```{r, echo=FALSE}
qplot(log(data$area), log(data$population), xlab ="log(area)", ylab = "log(population)")
```

## Step 3: Data aggregation by region
The areas and populations of all countries in a region are summed up using the **aggregate()** function, respectively. Then the following two pie charts are created using the **pie()** function.

```{r, echo=FALSE}
data.agg.area = aggregate(data$area, by=list(data$region), FUN=sum)
pie(data.agg.area$x, labels=data.agg.area$Group.1, main = "Area of regions")
#solution to 'integer overflow' for population data found on talkstats.com
data.agg.pop = aggregate(data$population, by=list(data$region), FUN = function(x){sum(as.numeric(x))}) 
pie(data.agg.pop$x, labels=data.agg.pop$Group.1, main = "Population of regions")
```

## Step 4: Visualization of Population per Square Kilometer by Regions
A new data frame is created to contain the population per sq km of each region using the **data.frame()** function. The data frame is then sorted by population per sq km in decreasing order with the help of the **order()** function. Finally, the following bar plot is created using the **qplot()** function with **geom=“bar”**.

```{r, echo=FALSE}
data.new = data.frame(region=data.agg.area$Group.1, pop.by.sq.km=(data.agg.pop$x/data.agg.area$x))
#data.new$pop.by.sq.km[order(data.new$pop.by.sq.km, decreasing=TRUE)]
attach(data.new)
sorted.data = data.new[order(-pop.by.sq.km),]
detach(data.new)
qplot(data.new$region, data.new$pop.by.sq.km, geom="bar", stat="identity", xlab="Region", ylab="Population per sq km of regions") + theme(axis.text.x = element_text(angle = 60, hjust = 1))
```
